VarSight: Prioritizing Clinically Reported Variants with Binary Classification Algorithms

### 2019-03-13 paper/data_template.tex
updating with some feedback

Exomiser$^*$&
{% for LEVEL in LEVELS %}
    {{ '%0.1f (%0.2f)' % (np.median(EXOMISER.TEST_RANKINGS[LEVEL]), np.mean(EXOMISER.TEST_RANKINGS[LEVEL])) }}&
{% endfor %}
{% for LEVEL in LEVELS %}
    {{ '%d, %d, %d' % 
        (100.0*np.sum(np.array(EXOMISER.TEST_RANKINGS[LEVEL]) <= 1) / len(EXOMISER.TEST_RANKINGS[LEVEL]),
        100.0*np.sum(np.array(EXOMISER.TEST_RANKINGS[LEVEL]) <= 10) / len(EXOMISER.TEST_RANKINGS[LEVEL]),
        100.0*np.sum(np.array(EXOMISER.TEST_RANKINGS[LEVEL]) <= 20) / len(EXOMISER.TEST_RANKINGS[LEVEL]))
    }}
    {% if LEVEL != 'PATHOGENIC' %}
        &
    {% endif %}
{% endfor %}
\\ \hhline{=|=|=|=|=||=|=|=|=}

>>>>

{% for ext_label in EXT_LABELS %}
    {{ ext_label.replace('_', '\\_') }}&
    {% for LEVEL in LEVELS %}
        {{ '%0.1f (%0.2f)' % (np.median(EXTERNAL[ext_label].TEST_RANKINGS[LEVEL]), np.mean(EXTERNAL[ext_label].TEST_RANKINGS[LEVEL])) }}&
    {% endfor %}
    {% for LEVEL in LEVELS %}
        {{ '%d, %d, %d' % 
            (100.0*np.sum(np.array(EXTERNAL[ext_label].TEST_RANKINGS[LEVEL]) <= 1) / len(EXTERNAL[ext_label].TEST_RANKINGS[LEVEL]),
            100.0*np.sum(np.array(EXTERNAL[ext_label].TEST_RANKINGS[LEVEL]) <= 10) / len(EXTERNAL[ext_label].TEST_RANKINGS[LEVEL]),
            100.0*np.sum(np.array(EXTERNAL[ext_label].TEST_RANKINGS[LEVEL]) <= 20) / len(EXTERNAL[ext_label].TEST_RANKINGS[LEVEL]))
        }}
        {% if LEVEL != 'PATHOGENIC' %}
            &
        {% endif %}
    {% endfor %}
    \\
{% endfor %}
\hhline{=|=|=|=|=||=|=|=|=}

### 2019-03-13 paper/main.tex
removing date so it doesn't act crazy later
updating with some feedback

\newif\ifpreprint
\preprinttrue

@@

\begin{document}
\title{VarSight: Prioritizing Clinically Reported Variants with Binary Classification Algorithms}
\author[1]{James M. Holt}
\author[1]{Brandon Wilk}
\author[1]{Camille L. Birch}
\author[1]{Donna M. Brown}
\author[1]{Manavalan Gajapathy}
\author[1]{Alexander C. Moss}
\author[1]{Nadiya Sosonkina}
\author[1]{Melissa A. Wilk}
\author[1]{Julie A. Anderson}
\author[1]{Jeremy M. Harris}
\author[1]{Jacob M. Kelly}
\author[1]{Fariba Shaterferdosian}
\author[1]{Angelina E. Uno-Antonison}
\author[1]{Arthur Weborg}
\author[ ]{Undiagnosed Diseases Network}
\author[1]{Elizabeth A. Worthey}
\affil[1]{Software Development and Informatics, HudsonAlpha Institute for Biotechnology, Huntsville, 35806, USA}
\maketitle

@@

In general, commonly used pipelines exist for the steps from sequencing through variant calling \citep{rehm2013, cornish2015}.  Despite differences in performance, most of these standards ingest the same information to create a list of variants from sequencing data.  In contrast, methods for variant annotation and/or variant filtering are quite diverse \citep{wang2010, hu2013, jager2014, desvignes2018}.  These methods use a wide range of input sources including but not limited to population allele frequencies \citep{lek2016}, conservation scores \citep{cooper2005, siepel2006, petrovski2013}, haploinsufficiency scores \citep{huang2010, steinberg2015}, deleteriousness scores \citep{steinberg2015, rentzsch2018}, transcript impact scores \citep{kumar2009, choi2012, adzhubei2013, dong2014, jian2014} , and previously associated disease annotation \citep{stenson2003, hamosh2005, landrum2015}.  Variant prioritization is also quite diverse with some methods relying on the variant annotations to prioritize variants \citep{hu2013} and some relying on patient phenotype to rank the variants \citep{kohler2009, yang2015, rao2018, wilk2018}.  There are also methods which combine both variant annotations and phenotype score to rank the variants \citep{singleton2014, zemojtel2014, smedley2015a}, a selection of which are benchmarked on the same simulated datasets in \cite{smedley2015b}.

@@

\subsection{Phenotype annotation}
The Codicem annotations are all agnostic of the patient phenotype.  As noted earlier, we expect these patient phenotypes to be noisy when compared to simulated datasets due to the variety and complexity of diseases, phenotypes, and genetic heritage tied to UDN patients.  In order to incorporate patient phenotype information, we used two distinct methods to rank genes based on the Human Phenotype Ontology (HPO) \citep{kohler2018}.  We then annotated each variant with the best scores from their corresponding gene(s).

The first method uses base annotations provided by the HPO to calculate a simple cosine score \citep{kohler2017} between the patient's phenotypes and each gene.  This method tends to be more conservative because it relies solely on curated annotations from the HPO.  The second method, an internally-developed tool called PyxisMap \citep{wilk2018}, uses the same annotations from the HPO, but adds in automatically text-mined data from NCBI's PubTator \citep{wei2013} and performs a Random-Walk with Restart \citep{page1999} on the ontology graph structure.  The PyxisMap method has the added benefit of incorporating gene-phenotype connections from recent papers that have not been manually curated into the HPO, but it also tends to make more spurious connections due to the imprecision of the text-mining from PubTator.  We used PyxisMap v1.2, and we ran the standard installation script that downloads all required data sources on December 19, 2018.  Each method generates a single numerical feature that is used in the following analyses.

@@

We then selected four readily available models for classification that are capable of training on imbalanced datasets: a random forest model by {\it sklearn} \citep{pedregosa2011}, a logistic regression model by {\it sklearn}, a balanced random forest model by {\it imblearn} \citep{lemaitre2017}, and an ensemble classification model by {\it imblearn}.  For each model, we selected a list of hyperparameters to test and tested each possible combination of those hyperparameters.  For each model and set of hyperparameters, we performed 10-fold cross validation on the training variants and recorded the balanced accuracy scores (i.e. a weighted accuracy score where inputs are weighted by their inverse class frequency).  For each model type, we saved the hyperparameters and model with the best average balanced accuracy score.  These four tuned models were then tested against the unseen set of test proband cases.
\ifpreprint

@@

For each tuned model, we calculated the 10-fold cross validated balanced accuracy on the training set.  We then calculated the true positive rate (TPR), false positive rate (FPR), and area under the receiver operator curve (AUROC) based on the unseen test data (see Table \ref{tab:classifier_results}).  
%The values for each model are shown in Table \ref{tab:classifier_results}.  
Figure \ref{fig:roc} show receiver operator curves for FPR v. TPR and recall v. precision for the four models.  %Figure \ref{fig:roc} also shows the receiver operator curves for two single-measure statistics (CADD Scaled and HPO-cosine) that are used in the next sections as comparisons for ranking variants.

From these metrics, the two random forest models and the EasyEnsembleClassifier have similar performance whereas whereas LogisticRegression performs slightly worse across the board.  
%All four classifiers have higher AUROC than either single-measure statistic.  
However, these classifier all have relatively poor performance from a precision-recall perspective (best AUROC for precision-recall was 0.2109).  This indicates that from a classification perspective, these models would identify a high number of false positives relative to the true positives unless a very conservative cutoff score was used.

\ifpreprint
\begin{table}[!t]
\centering
\begin{tabular}{c|c|c|c|c}
\hline
\input{classifier_rendered.tex}
\hline
\end{tabular}
\caption{Classifier performance statistics.  For each tuned classifier, we show performance measures commonly used for classifiers (from left to right): 10-fold cross validation balanced accuracy (CV10 Acc.), true positive rate (TPR), false positive rate (FPR), and area under the receiver operator curve (AUROC). In general, the two random forest methods and the EasyEnsembleClassifier perform very similarly whereas LogisticRegression seems to be slightly worse across all measures.}
\label{tab:classifier_results}
\end{table}

@@

\begin{table}[!t]
\processtable{Classifier performance statistics.  For each tuned classifier, we show performance measures commonly used for classifiers (from left to right): 10-fold cross validation balanced accuracy (CV10 Acc.), true positive rate (TPR), false positive rate (FPR), and area under the receiver operator curve (AUROC). In general, the two random forest methods and the EasyEnsembleClassifier perform very similarly whereas LogisticRegression seems to be slightly worse across all measures. \label{tab:classifier_results}}{
\begin{tabular}{c|c|c|c|c}
\hline
\input{classifier_rendered.tex}
\hline
\end{tabular}}{}
\end{table}
\fi

\ifpreprint
\newgeometry{margin=1cm} % modify this if you need even more space
\begin{landscape}
\begin{table}[!t]
\centering
\begin{tabular}{c|c|c|c|c||c|c|c|c}
\hline
\input{data_rendered.tex}
\hline
\end{tabular}
\caption{Ranking performance statistics.  This table shows the ranking performance statistics for all methods evaluated on our test set.  CADD Scaled and HPO-cosine are single value measures that were used as inputs to the classifiers we tested.  Exomiser is an external tool that only reported ranks for a subset of the filtered variants.  ``Exomiser(best)" conservatively assumed unranked primary variants were ranked at the next best position despite being unranked.  In contrast, ``Exomiser(avg)" realistically assumed unranked primary variants were at the average rank (i.e. middle) for all unranked variants.  The bottom four rows are the tuned, binary classification methods tested in this paper.  The ``Case Rank" columns show the median and mean ranks for all reported variants along with the variants split into their reported pathogenicity (calculated using the ACMG guidelines).  The ``Percentage in Top X Variants" columns show the percentage of variants that were found in the top 1, 10, and 20 variants in a case after ranking by the corresponding method.  All values were generated using only the test data that was unseen during training.}
\label{tab:performance}
\end{table}
\end{landscape}
\restoregeometry
\subsection{Ranking Statistics}

@@

\subsection{Ranking Statistics}
\begin{table*}[!t]
\centering
\processtable{Ranking performance statistics.  This table shows the ranking performance statistics for all methods evaluated on our test set.  CADD Scaled and HPO-cosine are single value measures that were used as inputs to the classifiers we tested.  Exomiser is an external tool that only reported ranks for a subset of the filtered variants.  ``Exomiser(best)" conservatively assumed unranked primary variants were ranked at the next best position despite being unranked.  In contrast, ``Exomiser(avg)" realistically assumed unranked primary variants were at the average rank (i.e. middle) for all unranked variants.  The bottom four rows are the tuned, binary classification methods tested in this paper.  The ``Case Rank" columns show the median and mean ranks for all reported variants along with the variants split into their reported pathogenicity (calculated using the ACMG guidelines).  The ``Percentage in Top X Variants" columns show the percentage of variants that were found in the top 1, 10, and 20 variants in a case after ranking by the corresponding method.  All values were generated using only the test data that was unseen during training. \label{tab:performance}}{
\begin{tabular}{c|c|c|c|c||c|c|c|c}
\hline
\input{data_rendered.tex}
\hline
\end{tabular}}{}%{$^*$Unranked variants were conservatively assumed to be at the next best rank after those that were ranked.}
\end{table*}
\fi

@@

For comparison, we selected to run Exomiser \citep{smedley2015a} because it performed comparatively well with all phenotypes even in the presence of noise (see benchmarking from \citealp{smedley2015b}).  We followed the installation on their website to install Exomiser CLI v.11.0.0 along with version 1811 for hg19 data sources.  For each test case, we created a VCF file from the pre-filtered list of annotated variants created by Codicem and passed those VCFs along with the full list of patient HPO terms to Exomiser.  We then parsed the output JSON file created by Exomiser into a ranked order of variants.

Unfortunately, Exomiser did not rank every variant that was present in the filtered VCF, typically ranking only 10-50 variants from the pre-filtered VCF.  For the reported variants, Exomiser did not rank 98 of the 189 variants in the test set (58 variant of uncertain significance, 21 likely pathogenic, and 19 pathogenic).  For these unranked variants, we assigned a rank using two different methods.  The ``Exomiser(best)" method conservatively assumed that unranked primary variants were found at the next best position in the rank order (e.g. if Exomiser ranked 15 variants, we assumed all unranked variants were ranked 16th). In contrast, the ``Exomiser(avg)" method assumed that unranked primary variants were ranked at the average position for all unranked variants (e.g. if Exomiser ranked 10 of 100 variants, we assumed all unranked variants were ranked 55th).  ``Exomiser(best)" represents the best possible ordering that Exomiser could give us for unranked variants where ``Exomiser(avg)" is a more realistic expectation.

Finally, we added two control scores for comparison: CADD scaled and HPO-cosine.  These scores were inputs to each classifier, but also represent two common ways one might naively order variants after filtering (by predicted deleteriousness and by similarity to phenotype).  The results for the two control scores, both Exomiser approaches, and all four classifiers are shown in Table \ref{tab:performance}.

In the overall data, all four classifiers outperform the single-measure statistics and both Exomiser approaches across the board, with the one exception being the mean of ``Exomiser(best)" (a measure that is heavily biased by the conservative ranking of unranked variants).  As one would intuitively expect, all classifiers perform better as the returned pathogenicity increases with the EasyEnsembleClassifier ranking 36\% of pathogenic variants in the first position and 97\% of pathogenic variants in the top 20.  Of the classifiers tested, the trained EasyEnsembleClassifier performs best overall with the strongest comparative performance in the pathogenic category.

\subsection{Random Forest Feature Importances}
After training and testing each classifier, we wished to explore which features played the largest role in how the classifier functioned.  Both random forest models calculate a feature importance array that stores how important each feature is in the trained model (the other two classifiers do not have this information readily available).  The total feature importance array sums to 1.0, and higher values indicate that a feature was used more by the trained model.  By chance, we had exactly 50 parameters as input to our model, thus we would expect approximately 2\% weight to be assigned to each parameter by chance.  Table \ref{tab:features} shows each features with an average importance $\ge 0.02$ across the random forest classifiers.

Interestingly, the two strongest individual features by far were the phenotype-based measures (HPO-cosine and PyxisMap) making up $\sim$33-40\% of the feature importance.  These were followed by CADD Scaled ($\sim$7-10\%) and three conservation scores ($\sim$14\%).  GnomAD-related fields ($\sim$8\%), GERP rsScore ($\sim$2-3\%), and HGMD-related fields ($\sim$4\%) made up the rest of features with higher than expected importance.  These results would suggest that phenotype-based metrics, deleteriousness scores, and conservation scores are of very high importance for predicting if a variant will be clinically returned to a rare disease patient.

\ifpreprint
\begin{table}[!t]
\centering
\begin{tabular}{l|c|c}
\hline
\input{features_rendered.tex}
\hline
\end{tabular}
\caption{Random forest feature importances.  This table shows the top feature importances reported by the two random forest algorithms we tested.  We had exactly 50 features passed into the classifier algorithms, so we show all features with an average importance $\ge 0.02$ indicating a feature utilized more than we would expect by chance.}
\label{tab:features}
\end{table}

@@

@@

@@

>>>>

\newif\ifpreprint
%\preprinttrue

@@

\begin{document}
\title{VarSight: Prioritizing Clinically Reported Variants with Binary Classification Algorithms}
\author[1]{James M. Holt}
\author[1]{Brandon Wilk}
\author[1]{Camille L. Birch}
\author[1]{Donna M. Brown}
\author[1]{Manavalan Gajapathy}
\author[1]{Alexander C. Moss}
\author[1]{Nadiya Sosonkina}
\author[1]{Melissa A. Wilk}
\author[1]{Julie A. Anderson}
\author[1]{Jeremy M. Harris}
\author[1]{Jacob M. Kelly}
\author[1]{Fariba Shaterferdosian}
\author[1]{Angelina E. Uno-Antonison}
\author[1]{Arthur Weborg}
\author[ ]{Undiagnosed Diseases Network}
\author[1]{Elizabeth A. Worthey}
\affil[1]{Software Development and Informatics, HudsonAlpha Institute for Biotechnology, Huntsville, 35806, USA}
\date{}
\maketitle

@@

In general, commonly used pipelines exist for the steps from sequencing through variant calling \citep{rehm2013, cornish2015}.  Despite differences in performance, most of these standards ingest the same information to create a list of variants from sequencing data.  In contrast, methods for variant annotation and/or variant filtering are quite diverse \citep{wang2010, hu2013, jager2014, desvignes2018}.  These methods use a wide range of input sources including but not limited to population allele frequencies \citep{lek2016}, conservation scores \citep{cooper2005, siepel2006, petrovski2013}, haploinsufficiency scores \citep{huang2010, steinberg2015}, deleteriousness scores \citep{steinberg2015, rentzsch2018}, transcript impact scores \citep{kumar2009, choi2012, adzhubei2013, dong2014, jian2014} , and previously associated disease annotation \citep{stenson2003, hamosh2005, landrum2015}.  Variant prioritization is also quite diverse with some methods relying on the variant annotations to prioritize variants \citep{hu2013} and some relying on patient phenotype to rank the variants \citep{kohler2009, yang2015, rao2018, wilk2018}.  There are also methods which combine both variant annotations and phenotype score to rank the variants \citep{singleton2014, zemojtel2014, smedley2015a, boudellioua2019}, a selection of which are benchmarked on the same simulated datasets in \cite{smedley2015b}.

@@

\subsection{Phenotype annotation}
The Codicem annotations are all agnostic of the patient phenotype.  As noted earlier, we expect these patient phenotypes to be noisy when compared to simulated datasets due to the variety and complexity of diseases, phenotypes, and genetic heritage tied to UDN patients.  Additionally, we made no effort to alter or condense the set of phenotypes provided by the corresponding clinical sites.  In order to incorporate patient phenotype information, we used two distinct methods to rank genes based on the Human Phenotype Ontology (HPO) \citep{kohler2018}.  We then annotated each variant with the best scores from their corresponding gene(s).

The first method uses phenotype-to-gene annotations provided by the HPO to calculate a cosine score \citep{kohler2017} between the patient's phenotypes and each gene.  In short, this method builds a vector from the phenotype terms (and ancestor terms in the ontology) for both the patient and each gene.  Then, the cosine between the patient vector and each gene vector is calculated as a representation of similarity.  This method tends to be more conservative because it relies solely on curated annotations from the HPO.

The second method, an internally-developed tool called PyxisMap \citep{wilk2018}, uses the same phenotype-to-gene annotations from the HPO, but adds in automatically text-mined annotations from NCBI's PubTator \citep{wei2013} and performs a Random-Walk with Restart \citep{page1999} on the ontology graph structure.  The PyxisMap method has the added benefit of incorporating gene-phenotype connections from recent papers that have not been manually curated into the HPO, but it also tends to make more spurious connections due to the imprecision of the text-mining from PubTator.  We used PyxisMap v1.2, and we ran the standard installation script that downloads all required data sources on December 19, 2018.  Each method generates a single numerical feature that is used in the following analyses.

@@

We then selected four publicly available models for classification that are capable of training on imbalanced datasets: a random forest model by {\it sklearn} \citep{pedregosa2011}, a logistic regression model by {\it sklearn}, a balanced random forest model by {\it imblearn} \citep{lemaitre2017}, and an ensemble classification model by {\it imblearn}.  The primary reason for selecting these four classifiers is that they are all capable of handling imbalanced data either by default or with a single parameter specifying to balance the data during training.  The two random forest models build collections of decision trees that weight each training input by its class frequency.  Logistic regression calculates the probability of a value belonging to a particular class, again weighting by the class frequency.  In contrast to the other three tested methods, the ensemble classification balances the training input using random under-sampling and then trains an ensemble of AdaBoost learners.

Initially, we also tried the support vector classifier by {\it sklearn}, the multi-layer perceptron by {\it sklearn}, and the random under-sampling AdaBoost classifier by {\it imblearn}.  Each of these was excluded from our results due to, respectively, scaling issues with the training size, failure to handle the data imbalance, and overfitting to the training set.  While we did not achieve good results using these out-of-the-box implementations, it may be possible to use the methods with appropriate changes made to the underlying implementations.

For each of our tested models, we selected a list of hyperparameters to test and tested each possible combination of those hyperparameters.  For each model and set of hyperparameters, we performed 10-fold cross validation on the training variants and recorded the balanced accuracy scores (i.e. a weighted accuracy score where inputs are weighted by their inverse class frequency).  For each model type, we saved the hyperparameters and model with the best average balanced accuracy score.  These four tuned models were then tested against the unseen set of test proband cases.
\ifpreprint

@@

For each tuned model, we calculated the 10-fold cross validated balanced accuracy on the training set.  We then calculated the area under the receiver operator curve (AUROC) and area under the precision-recall curve (AUPRC) using the unseen test set (see Table \ref{tab:classifier_results}).  Figure \ref{fig:roc} shows the corresponding receiver operator curves and precision-recall curves for the four models.

From these metrics, the two random forest models and the EasyEnsembleClassifier have similar performance whereas whereas LogisticRegression performs slightly worse across the board.  However, these classifier all have relatively poor performance from a precision-recall perspective (best AUROC for precision-recall was 0.2109).  This indicates that from a classification perspective, these models would identify a high number of false positives relative to the true positives unless a very conservative cutoff score was used.

\ifpreprint
\begin{table}[!t]
\centering
\begin{tabular}{c|c|c|c}
\hline
\input{classifier_rendered.tex}
\hline
\end{tabular}
\caption{Classifier performance statistics.  For each tuned classifier, we show performance measures commonly used for classifiers (from left to right): 10-fold cross validation balanced accuracy (CV10 Acc.), area under the receiver operator curve (AUROC), and area under the precision-recall curve (AUPRC). In general, the two random forest methods and the EasyEnsembleClassifier perform very similarly whereas LogisticRegression seems to be slightly worse across all measures.}
\label{tab:classifier_results}
\end{table}

@@

\begin{table}[!t]
\processtable{Classifier performance statistics.  For each tuned classifier, we show performance measures commonly used for classifiers (from left to right): 10-fold cross validation balanced accuracy (CV10 Acc.), area under the receiver operator curve (AUROC), and area under the precision-recall curve (AUPRC). In general, the two random forest methods and the EasyEnsembleClassifier perform very similarly whereas LogisticRegression seems to be slightly worse across all measures. \label{tab:classifier_results}}{
\begin{tabular}{c|c|c|c}
\hline
\input{classifier_rendered.tex}
\hline
\end{tabular}}{}
\end{table}
\fi

\ifpreprint
\newgeometry{margin=1cm} % modify this if you need even more space
\begin{landscape}
\begin{table}[!t]
\centering
\begin{tabular}{c|c|c|c|c||c|c|c|c}
\hline
\input{data_rendered.tex}
\hline
\end{tabular}
\caption{Ranking performance statistics.  This table shows the ranking performance statistics for all methods evaluated on our test set.  CADD Scaled and HPO-cosine are single value measures that were used as inputs to the classifiers we tested.  
%Exomiser is an external tool that only reported ranks for a subset of the filtered variants.  ``Exomiser(best)" conservatively assumed unranked primary variants were ranked at the next best position despite being unranked.  In contrast, ``Exomiser(avg)" realistically assumed unranked primary variants were at the average rank (i.e. middle) for all unranked variants.  
The middle four rows (two Exomiser runs, Phen-Gen, and DeepPVP) represent external tools that ranked the same set of variants as the classifier algorithms.  Phen-Gen was the only external tool that did not rank every variant in the set, so we conservatively assumed unranked variants were at the next best position despite being unranked.
The bottom four rows are the tuned, binary classification methods tested in this paper.  The ``Case Rank" columns show the median and mean ranks for all reported variants along with the variants split into their reported pathogenicity (calculated using the ACMG guidelines).  The ``Percentage in Top X Variants" columns show the percentage of variants that were found in the top 1, 10, and 20 variants in a case after ranking by the corresponding method.  All values in this table were generated using only the test data that was unseen during training.}
\label{tab:performance}
\end{table}
\end{landscape}
\restoregeometry
\subsection{Ranking Statistics}

@@

\subsection{Ranking Statistics}
\begin{table*}[!t]
\centering
\processtable{Ranking performance statistics.  This table shows the ranking performance statistics for all methods evaluated on our test set.  CADD Scaled and HPO-cosine are single value measures that were used as inputs to the classifiers we tested.  
%Exomiser is an external tool that only reported ranks for a subset of the filtered variants.  ``Exomiser(best)" conservatively assumed unranked primary variants were ranked at the next best position despite being unranked.  In contrast, ``Exomiser(avg)" realistically assumed unranked primary variants were at the average rank (i.e. middle) for all unranked variants.  
The middle four rows (two Exomiser runs, Phen-Gen, and DeepPVP) represent external tools that ranked the same set of variants as the classifier algorithms.  Phen-Gen was the only external tool that did not rank every variant in the set, so we conservatively assumed unranked variants were at the next best position despite being unranked.
The bottom four rows are the tuned, binary classification methods tested in this paper.  The ``Case Rank" columns show the median and mean ranks for all reported variants along with the variants split into their reported pathogenicity (calculated using the ACMG guidelines).  The ``Percentage in Top X Variants" columns show the percentage of variants that were found in the top 1, 10, and 20 variants in a case after ranking by the corresponding method.  All values in this table were generated using only the pre-filtered variants from testing data. \label{tab:performance}}{
\begin{tabular}{c|c|c|c|c||c|c|c|c}
\hline
\input{data_rendered.tex}
\hline
\end{tabular}}{}%{$^*$Unranked variants were conservatively assumed to be at the next best rank after those that were ranked.}
\end{table*}
\fi

@@

For comparison, we selected to run Exomiser \citep{smedley2015a}, Phen-Gen \citep{javed2014}, and DeepPVP \citep{boudellioua2019}.  For each tool, we input the exact same set of phenotype terms used by the classifiers we tested.  Additionally, we used the pre-filtered variant set from Codicem as input to each ranking algorithm.  As a result, all external tools and our trained classifiers are ranking on identical phenotype and variant information.

For Exomiser, we followed the installation on their website to install Exomiser CLI v.11.0.0 along with version 1811 for hg19 data sources.  We ran Exomiser twice, once using the default hiPhive prioritizer (incorporates knowledge from human, mouse, and fish) and once using the human only version of the hiPhive prioritizer (this was recommended in place of the deprecated PhenIX algorithm \citep{zemojtel2014}).  Phen-Gen V1 was run using the pre-compiled binary using the ``dominant'' and ``genomic'' modes to maximize the output.  Of note, Phen-Gen was the only external method that did not fully rank all variants, so we conservatively assumed that any absent variants were at the next best possible rank.  Thus, the reported Phen-Gen comparisons are an optimistic representation for this test data.  Finally, DeepPVP v2.1 was run using the instructions available on their website.  Details on the exact execution for each external tool can be found in the Supplementary Documents (TODO: XXX).

Finally, we added two control scores for comparison: CADD scaled and HPO-cosine.  These scores were inputs to each classifier, but also represent two common ways one might naively order variants after filtering (by predicted deleteriousness and by similarity to phenotype).  The results for the two control scores, all four external tools, and all four trained classifiers are shown in Table \ref{tab:performance}.

In the overall data, all four classifiers outperform the single-value measures and external tools across the board.  As one would intuitively expect, all classifiers perform better as the returned pathogenicity increases with the EasyEnsembleClassifier ranking 36\% of pathogenic variants in the first position and 97\% of pathogenic variants in the top 20.  Of the classifiers tested, the trained EasyEnsembleClassifier performs best overall with the strongest comparative performance in the pathogenic category.

\subsection{Random Forest Feature Importances}
\ifpreprint
\begin{table}[!t]
\centering
\begin{tabular}{l|c|c}
\hline
\input{features_rendered.tex}
\hline
\end{tabular}
\caption{Random forest feature importances.  This table shows the top feature importances reported by the two random forest algorithms we tested.  We had exactly 50 features passed into the classifier algorithms, so we show all features with an average importance $\ge 0.02$ indicating a feature utilized more than we would expect by chance.}
\label{tab:features}
\end{table}

@@

After training and testing each classifier, we wished to explore which features played the largest role in how the classifier functioned.  Both random forest models calculate a feature importance array that stores how important each feature is in the trained model (the other two classifiers do not have this information readily available).  The total feature importance array sums to 1.0, and higher values indicate that a feature was used more by the trained model.  By chance, we had exactly 50 parameters as input to our model, thus we would expect approximately 2\% weight to be assigned to each parameter by chance.  Table \ref{tab:features} shows each features with an average importance $\ge 0.02$ across the random forest classifiers.

The two strongest individual features by far were the phenotype-based measures (HPO-cosine and PyxisMap) making up $\sim$33-40\% of the feature importance.  This is not a major surprise given that phenotype information was not used during the pre-filtering steps provided by Codicem.  These were followed by CADD Scaled ($\sim$7-10\%) and three conservation scores ($\sim$14\%).  These results would suggest that phenotype-based metrics, deleteriousness scores, and conservation scores are of very high importance for the classifiers we tested.

@@

%DeepPVP - variant ranker
\bibitem[Boudellioua {\it et~al}., 2019]{boudellioua2019} Boudellioua, Imane, \textit{et~al.} "DeepPVP: phenotype-based prioritization of causative variants using deep learning." {\it BMC bioinformatics} 20.1 (2019): 65.

@@

%Phen-Gen - variant ranker
\bibitem[Javed {\it et~al}., 2014]{javed2014} Javed, Asif, Saloni Agrawal, and Pauline C. Ng. "Phen-Gen: combining phenotype and genotype to analyze rare disorders." {\it Nature methods} 11.9 (2014): 935.
